### 2017-07-03 chapters/introduction/patient.tex
Update latest

In order to make use of models

\subsection{Data aquistition}

@@

The interaction between cardiac images and cardiac models are important.

Still, one of the major bottlenecks in terms of personalized
computaional models is the quality of data (and representations of
model). There is a big issue in
reproducebility, because of noisy data, and operator depedent
measurements.

Today there are three main non-invasive imaging techniques used in
cardiology which is echocardiography (ECHO), magnetic
resonance imaging (MRI) and computed tomography (CT), and all of them
are used for aquiring data used in computaional modeling. Each
modality offers advantages and disadvatages over the other, and we
will make a short summary of pros and cons for each modelity below.

\subsubsection{CT}
Computed tomography is reconginzed as the

\subsubsection{ECHO}

\subsubsection{MRI}

@@

The main modality used in this thesis is 4D echocardiography, and we
will therefore focus on data aquired using this modality.
With 4D we mean three spatial and one temporal dimesion.
In order to aquire a 3D volume image using ultrasound, the speed of sound in
human tissue (which is approxmately 1540 m/s) put some limitations on
the image quality verus the frame-rate \cite{rabben2010technical}.
Consequently, a 3D ultracound image is formed by stiching together $N$
disjoint subvolumes aquired during $N$ cardiac cycles ($N$ typicall
between 2 and 8)\cite{brekke2007volume}.

The images are later processed using some image segmentation tool.
One such tool is EchoPac which is used for analysing images for GE
Vingmed. The 4D Auto LVQ tool in EchoPac, is a tool for processing 3D
echo images, and traces of volume, triangulated surfaces and 3D strain
traces together with a structured mesh of the AHA segments for each
time point, can be exported to and HDF format and later used for mesh generation.

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/raw.png}
    \caption{\label{fig:echopac_out_surf}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh.png}
    \caption{\label{fig:echopac_out_strain_mesh}}
  \end{subfigure}
\caption{}
\label{fig:echopac_output}
\end{figure}

As discussed in \ref{sec:mech_boudary}, we constrain the basal movement in
the longitudinal direction, which is easiest to accomplished using a flat
base located at a prescribed location. In order to make the base flat
we first orient the surfaces so that the longitudinal axis is aligned
with the $x$-axis, and the apex pointing in the postive $x$
direction. To conststruct the basal plane we first take out the basal
points from the strain mesh, and use these points to construct a least
sqauare fitting plane (Figure \ref{fig:strain_mesh_plane}). Let $(x_i, y_i, z_i), i = 1, \cdots, N$
be the basal points, and suppose the basal plane solves the equation
$z = ax + by + c$, for some unknown constants $a,b,c$. Following a
least square approach, we select the parameters $(a,b,c)$ that
minimizes to sum
\begin{align}
  \sum_{i = 1}^{N} \left( z_i - ax_i + by_i + c \right)^2.
\end{align}
Once the parameters for the basal plane is found we adjust the size of
the cut, by moving the plane along the longitudinal axis (here
$x-$axis), until the cavity volume agrees with the measured volume
given by the image segmentation tool within some specified tolerance. 
When the correct size is found the points above the plane is removed,
and the remaining surfaces are smoothed using the GAMer
\cite{yu2008feature}.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh_plane.pdf}
    \caption{\label{fig:strain_mesh_plane}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/cut.png}
    \caption{\label{fig:cut}}
  \end{subfigure}
\caption{}
\label{fig:echopac_output}
\end{figure}

The actualy mesh generation is performed using Gmsh
\cite{geuzaine2009gmsh}, which meshes the endocardial and epicardial
surface togther unsing frontal-Delaunay meshing algorithm. Gmsh also marks
the endocardal, the epicardial  and the basal facets, along with the
endocardial

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/mesh.png}
    \caption{\label{fig:mesh}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/markers_full.pdf}
    \caption{\label{fig:markers_full}}
  \end{subfigure}
\caption{}
\label{fig:echopac_output}
\end{figure}

@@

\subsubsection{Rule-based fibers}
\label{sec:rule_based_fiber}
Mention papers about fiber distribution, fiber dependence and
different rule-based fibers.

Describe the main steps in the Bayer algorithm.

@@

\begin{align}
  \begin{split}
    x &= a \sinh \mu \sin \nu \cos \theta  \\
    y &= a \sinh \mu \sin \nu \sin \theta  \\
    z &= a \cosh \mu \cos \nu,
  \end{split}
\end{align}
where $a$ is the focal point, $\nu \in [0, \pi]$ is the longitudinal
coordinate, $\theta \in [0, 2\pi]$ is the azimuthal angle and $\mu$
is the radial coordinate. For an ellipse given by the equation
$\frac{x^2}{b^2} +\frac{y^2}{c^2}  = 1$ with $b$ being the semimajor
axis, and $c$ the semiminor axis, the focal point is given by
\begin{align}
  a =\sqrt{b^2 - c^2}.
  \label{eq:focal_point}
\end{align}
Assuming the ventricle is axisymetric around the
longitudinal axis, which is aligned with the $x$-axis it is possible
to estimate the focal point by taking $b$ in \eqref{eq:focal_point} to
be the maximum distance from the base to the apex, and $b$ the maximum
radius at the based. Using this estimate, the base is assumed to be
located approximately at the center of the ellipsoid, i.e $\nu = \pi /
2$. Compared to geometries used in benchmarking of cardiac mechanics
problems, this is not a totally wrong assumption
\cite{land2015verification}

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate.png}
    \caption{\label{fig:prolate_coord}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate_cube.png}
    \caption{\label{fig:prolate_cube}}
  \end{subfigure}
\caption{}
\label{fig:prolate}
\end{figure}

\subsubsection{Marking the mesh according to the AHA-segments}

\begin{itemize}
  \item Segmentation and aquisition using echopach
  \item From closed endo- and epicardial surfaces to a mesh
    (smoothing, cutting, orienting).
  \item Marking of mesh accoring to AHA segments (16,17,18)
  \item Local basis
  \item Rule based fibers
  \item Referece to mesh{\_}generation and patient repositories
  \end{itemize}

@@

The other approach called variational data assimilation, which is the
one used in this thesis, is based on minimizing a cost functional that
represents the mismatch between observations and simulations, with
possibly additional regularization terms added to the objective
functional.

% The general setup for a variational data assimilation problem is the
% following. 
We will now explain the general setup for variational data
assimilation. In order to keep consistent notation we will refer to
the \emph{state variable} as $\mathbf{w}$, which is our case
represents jointly the displacement and hydrostatic pressure
$\mathbf{w}=(\mathbf{u}, p)$.

The state varibles are described by some \emph{phyiscal
model} $\mathcal{L}$, which is our case is governed by the force balance
equations \eqref{eq:force_balance}, and can be written in the form
$\mathcal{L}\mathbf{w} = f$.

We are also given a set of measurements (or observations) that we want
to assimilate, and we denote this by the vector $\mathbf{y}$.
For example, $\mathbf{y}$ might be a set of volume measurements or a
combinations of volume and strain at given time points in the cardiac
cycle. The \emph{observation operator} $\mathcal{H}$, is approximation
of the observation and act as an operator from the state space to the
observation space. We therefore have the relation
\begin{align}
  \mathbf{y} = \mathcal{H}(\mathbf{w}) + \xi_o, 
\end{align}
where $\xi_o$ represents both the error in the measurements (e.g
noise in the data) and error in the representation of the observation
(that $\mathcal{H}$ do not represent the data good enough).

\paragraph{The volume observation operator
  $\mathcal{H}_{\mathrm{volume}}$}
In some cases, the ventricular cavity volume can easily be computed
analytically from the displacement field. Denote the inner cavity of
the left ventricle in the current configuration by
$\omega_{\mathrm{endo}}$. Further let $\mathrm{d} v$ and $\xvec$
denote an infinitesimal volume element and the coordinate in the
current configuration respectively. Then by the divergece theorem we have
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= \int_{\omega_{\mathrm{endo}}} \mathrm{d} v =
  \frac{1}{3}\int_{\omega_{\mathrm{endo}}} \nabla \cdot \xvec \mathrm{d} v =
  \frac{1}{3}\int_{\partial \omega_{\mathrm{endo}}} \xvec \cdot \mathbf{n} \mathrm{d} s, 
\end{align}
where $\mathbf{n}$ and $\mathrm{d} s$ are the unit normal and a surface
element on the current configuration respectively.
Note that the boundary $\partial \omega_{\mathrm{endo}}$ includes the
endocardial basal boundary. However in the case when the base is flat
and fixed in longitudinal direction, the contribution from this
boundary integral will be zero. In this case the boundary integral
will be the same as the integral over the endocardial boundary
decribed in \ref{refer_to_section}, with the only change being the
change of sign on the normal vector. By Nansens formula we obtain
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= -\frac{1}{3}\int_{\lvendo} \left( \Xvec + \uvec \right) J \F^{-T} \Nvec \mathrm{d}S,
  \label{eq:volume_operator}
\end{align}
where now $\lvendo$,  $\Xvec$, $\Nvec$ and $\mathrm{d}S$ are respectively the
endocardial surface, the referece coordinate, the unit normal and a
surface element in the reference configuration.

\paragraph{The strain observation operator
  $\mathcal{H}_{\mathrm{volume}}$}
Another observation that is encountered in this thesis is strain
measurements, of more precisely average regional strain in the
circumferential, radial or longitudinal direction. Let $\Omega_j$
denote the volume for which the strain should be averaged over, and
let $\mathbf{e}_k$ denote the unit vectorfield in the preferred strain
direction. Then for a given strain tensor $\mathbf{A}$ we define 
\begin{align}
  \mathcal{H}_{\mathrm{strain}}(\uvec) = \frac{1}{|\Omega_j|}\int_{\Omega_j} \mathbf{e}_k^T \mathbf{A}(\uvec) \mathbf{e}_k  \mathrm{d}V
\end{align}
The strain tensor $ \mathbf{A}$ is typically chosen to be the
Grenn-Lagrange strain tensor $\mathbf{E}$\eqref{eq:green_lagrange}, or
the material displacement gradient tensor $\F - \I$.

\begin{remark}
  Measured strains are computed relative to some reference
  geometry, which do not always coinside with the reference geometry
  chosen for your simulation. In these cases you have to either
  recompute the measured strain according to the chosen referece for
  the simulation, or recompute the simulated strains according to the
  correct referece geometry for your measuments. 
\end{remark}

% Within varitaional data assimilation it is common to separate between
% dynamic models and static models. These are conventionally reffered to as
% $4D-$Var and $3D-$Var methods respectively. In the work done this
% thesis we have used both of these methods.

% For an introduction to data assimilation we refer to e.g \cite{lahoz2010data}.

Let us assume that

\begin{itemize}
  \item Examples of data you want to constrain
  \item Forming of mismatch functional and problem formulation as
    PDE-constrained optimization.
  \item The adjoint approach. comparison with other methods.
  \item Control parameters
  \item OPtimization algorithm - stopping criteria
  \item Reference to pulse{\_}adjoint and pulse{\_}adjoint{\_}post
    repositories
  \item Variational versus sequnetial
  \item Question about uniquness
\end{itemize}

We will now explain what we mean by ``adjoint-based'', and why this
approach is a key ingredient. The main theory presented here is taken
from the dolfin-adjoint web page. A word about dolfin adjoint...\ref{}

\begin{equation}
  \begin{aligned}
    \label{eq:opt_matparam}
    & \underset{\mvec}{\text{minimize}}
    & &  \mathcal{J}(\state, \mvec) \\
    & \text{subject to}
    & & \delta \Pi(\state, \mvec) = 0, 
  \end{aligned}
\end{equation}

>>>>

In this section we will go through the steps needed to personalize the
mechanical models developed in the previous section towards a given
patient. We will start by giving a brief overview of the
interaction between medical imaging and cardiac computational model.
Then we will explain how the geometries are extracted from the
medical images and turned into a computational mesh,  which we can use
for finite element computations. Finally we will explain how to
incorporate additional data into model by means of adjoint based data
assimilation techniques.

\subsection{Medical Imaging}

@@

Cardiac computational modeling often relies on data obtained using
medical imaging techniques, and the quality of such data is therefore of
fundamental importance \cite{lamata2014images}. It is clear, that
without advancements in medical imaging technology, the role of
computational models would be limited to simple, idealized cases. With
access to high quality images, detailed information about cardiac structure
can be used to e.g build anatomically correct models. The
computational models can then later be validated agains this images,
or we can assimilate additional data extracted from this images to
extract biomarkers. Figure \ref{fig:cardiac_imaging_model} summarizes
this interaction between medical images and computational models.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/models.eps}
\caption{Interaction between medical imaging and computational
  models. Medical images are used to extract geometric
  structure. These geometries together with the computational models
  can later be validated against the medical images. Computational
  models can also be used togther with the medical images to
  assimilate additional data in order to extraction clinically useful
  biomarkers. }
\label{fig:cardiac_imaging_model}
\end{figure}

% Still, one of the major bottlenecks in terms of personalized
% computaional models is the quality of data (and representations of
% model). There is a big issue in
% reproducebility, because of noisy data, and operator depedent
% measurements.
Today there are three main non-invasive imaging techniques used in
cardiology. These are chocardiography (Echo), magnetic
resonance imaging (MRI) and computed tomography (CT). Each
modality offers advantages and disadvatages over the other

@@

% \subsubsection{EchoPac}

The main modality used in this thesis is 4D echocardiography, and we
will therefore focus on data aquired using this modality.
With 4D we mean three spatial and one temporal dimesion.
The speed of sound in human tissue (which is approxmately 1540 m/s)
put some limitations on the image quality verus the frame-rate
\cite{rabben2010technical} for these 3D volume images. In order to
increase the resolution, such an image is formed by stiching together
$N$ disjoint subvolumes aquired during $N$ cardiac cycles ($N$
typicall between 2 and 8)\cite{brekke2007volume}.

The images are later processed using some image segmentation tool.
One such tool is EchoPac which is used for analysing images for GE
Vingmed. The 4D Auto LVQ tool in EchoPac, is a tool for processing 3D
echo images, and traces of volume, triangulated surfaces and 3D strain
\cite{heimdal20114d} traces together with a structured mesh of the
American Heart Association (AHA) segments
\cite{cerqueira2002standardized} for each time point, see Figure
\ref{fig:echopac_output}.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/raw.png}
    \caption{\label{fig:echopac_out_surf}}
  \end{subfigure}
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh.png}
    \caption{\label{fig:echopac_out_strain_mesh}}
  \end{subfigure}
  \begin{subfigure}[t]{0.38\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/bullseye/bullseye.pdf}
    \caption{\label{fig:bullseye_intro}}
  \end{subfigure}
\caption{Figure \ref{fig:echopac_out_surf} shows triangulated surfaces of the
      endo- and epicardium, and Figure
      \ref{fig:echopac_out_strain_mesh} shows structured mesh of the
      American Heart Association (AHA) segments. These surfaces are
      exported from the image segmentational tool EchoPac. Figure 
      \ref{fig:bullseye_intro} shows the American Heart Association (AHA)
      segments illustrated on a so called bullseye plot with name and
      numbers on the different segments. }
\label{fig:echopac_output}
\end{figure}

@@

As discussed in \ref{sec:mech_boudary}, we constrain the basal movement in
the longitudinal direction, which is easiest to accomplished using a flat
base located at a prescribed location. In order to make the base flat
we first orient the surfaces so that the longitudinal axis is aligned
with the $x$-axis, and the apex pointing in the postive $x$
direction. To conststruct the basal plane we first take out the basal
points from the strain mesh, and use these points to construct a least
sqauare fitting plane (Figure \ref{fig:strain_mesh_plane}). Let $(x_i, y_i, z_i), i = 1, \cdots, N$
be the basal points, and suppose the basal plane solves the equation
$z = ax + by + c$, for some unknown constants $a,b,c$. Following a
least square approach, we select the parameters $(a,b,c)$ that
minimizes to sum
\begin{align}
  \sum_{i = 1}^{N} \left(ax_i + by_i + c  - z_i \right)^2.
\end{align}
Once the parameters for the basal plane is found we adjust the size of
the cut, by moving the plane along the longitudinal axis (here
$x-$axis), until the cavity volume agrees with the measured volume
given by the image segmentation tool within some specified tolerance. 
When the correct size is found the points above the plane is removed,
and the remaining surfaces are smoothed using the GAMer
\cite{yu2008feature}.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh_plane.pdf}
    \caption{\label{fig:strain_mesh_plane}}
  \end{subfigure}
  \begin{subfigure}[t]{0.36\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/cut.png}
    \caption{\label{fig:cut_intro}}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/mesh.png}
    \caption{\label{fig:mesh_intro}}
  \end{subfigure}
\caption{A least square fitting plane is fitted to the points
  belonging to the basal boundary in the strain mesh (Figure
  \ref{fig:strain_mesh_plane}). This plane is used to cut the endo- and
  epicardial surfaces (Figure \ref{fig:cut_intro}) and Gmsh is used to mesh
  these surfaces together (Figure \ref{fig:mesh_intro}). }
\label{fig:mesh_generation_intro}
\end{figure}

The actualy mesh generation is performed using Gmsh
\cite{geuzaine2009gmsh}, which meshes the endocardial and epicardial
surface togther unsing frontal-Delaunay meshing algorithm. Gmsh also marks
the endocardal, the epicardial  and the basal facets, along with the
endocardial and epicardial basal rings.

% \begin{figure}[htbp]
%   \centering

%   \begin{subfigure}[t]{0.45\textwidth}
%     \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/markers_full.pdf}
%     \caption{\label{fig:markers_full}}
%   \end{subfigure}
% \caption{}
% \label{fig:echopac_output}
% \end{figure}

@@

\subsubsection{Rule-based fiber architecture}
\label{sec:rule_based_fiber}
Both the passive and active properties of the myocardium depends on
the underlying muscle fiber architecture, and it is therefore of high
relevance to capture this architecture as accuratley as
possible. Unfortunately, imaging technology does not currently provide a
way of extracting patients-specific fiber ortiention \emph{in
  vivo}. Diffusion Tensor Imaging (DTI) can be utilized to reconstruct
the fiber and lamiar structure \cite{rohmer2007reconstruction}
\emph{ex vivo}.

An alternative method for assigning myocardial fiber orientation is by
rule-based methods \cite{potse2006comparison, bayer2012novel}. 
In this thesis we have used an algorithm called the Laplace–Dirichlet
Rule-Based (LDRB) algorithm, proposed by Bayer et al \cite{bayer2012novel}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]
  {chapters/introduction/figures/fiber/fiber}
\caption{The LDRB algorithm \cite{bayer2012novel} applied to an LV
  (left) and BiV (right) geometry with an helix fiber angle of
  $60^{\circ}$ and $-60^{\circ}$ on the endo- and epicardium
  respectively. Color represents the helical angle, with blue being
  $0^{\circ}$ (circumferential) which can be seen in the midwall. }
\label{fig:echopac_output}
\end{figure}

The LDRB algorithm takes as
input the geometry $\Omega$, the helical fiber angle on the endo- and
epicardium ($\alpha_{\mathrm{endo}}$ and $\alpha_{\mathrm{epi}}$) and
the transverse fiber angle on the endo- and epicardium
($\beta_{\mathrm{endo}}$ and $\beta_{\mathrm{epi}}$), and outputs a
set of three orthogonal vector fields, one representing the fiber
angle $(\ef)$, one representing the sheet angle $(\es)$ and the final
diraciton is called the sheet normal direction $(\en)$

@@

\begin{align}
  \begin{split}
    x &= a \sinh \mu \sin \nu \cos \theta  \\
    y &= a \sinh \mu \sin \nu \sin \theta  \\
    z &= a \cosh \mu \cos \nu,
  \end{split}
\end{align}
where $a$ is the focal point, $\nu \in [0, \pi]$ is the longitudinal
coordinate, $\theta \in [0, 2\pi]$ is the azimuthal angle and $\mu$
is the radial coordinate. For an ellipse given by the equation
$\frac{x^2}{b^2} +\frac{y^2}{c^2}  = 1$ with $b$ being the semimajor
axis, and $c$ the semiminor axis, the focal point is given by
\begin{align}
  a =\sqrt{b^2 - c^2}.
  \label{eq:focal_point}
\end{align}
The prolate spheroidal coordinate axes are illustrated in
Figure~\ref{fig:prolate_coord}. There inverse relation, going from the
cartesian coordinate $(x,y,z)$ to the prolate spheroidal coordinate
$(\mu, \nu, \theta)$ given the focal point $a$ is a bit more
intricate, but can be computed as follows 
\begin{align}
  \begin{split}
    \nu &= \arccos (\tau) \\
    \mu &= \arccosh (\sigma) \\
    \theta &= \arctan \frac{z}{y} \\
    \tau &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2} \\
    & \quad + \sqrt{(x - a)^2 + y^2 + z^2} \big) \\
    \sigma &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2}\\
    & \quad - \sqrt{(x - a)^2 + y^2 + z^2}\big) 
  \end{split}
\end{align}
It is possible to estimate the focal point by taking $b$ in
\eqref{eq:focal_point} to be the maximum distance from the base to the
apex, and $c$ the maximum radius at the base. Using this estimate,
the base is assumed to be located approximately at the center of the
ellipsoid, i.e $\nu = \pi / 2$ \cite{land2015verification}.
% Compared to geometries used in
% benchmarking of cardiac mechanics problems, this is not a totally
% wrong assumption 
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate.png}
    \caption{\label{fig:prolate_coord}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate_cube.png}
    \caption{\label{fig:prolate_cube}}
  \end{subfigure}
\caption{}
\label{fig:prolate}
\end{figure}

This coordinate system can be used when e.g marking the mesh according
to the AHA segments

\subsection{Choosing the reference geometry}
\label{sef:reference_geometry}
The choice of reference geometry ..
 choosing the geometry with the smallest
volume as the reference geometry (i.e the end-systolic geometry)
(references), the mid-diastolic or the end-diastolic ??

A general problem in the biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood pressure.
This is analogous to finding the geometry of a deflated balloon, given
an image of and inflated one. Having this picture in mind, it is
intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to obtain the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. We briefly
explain the steps in this methods.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.9\textwidth]{chapters/introduction/figures/unloading/canvas.pdf}
\caption{Showing 1, 5, and 10 iterations of the backward displacement
  method applied to a left ventricular geometry. To row top shows the
  original geometry in solid and the inflated unloaded geometry in
  yellow wire-frame for 1,5 and 10 iterations. Bottom row shows the
  unloaded geometry in solid and the original image based geometry in wire-frame.}
\label{fig:unloading_lv}
\end{figure}

Suppose you are given geometry $\Omega_{\mathcal{I}}$ taken from some medical image, and
is thus loaded with some loads $\mathbf{T}$, and we want to find the unloaded geometry
$\Omega_{\mathcal{U}}$, which is so that when we load $\Omega_{\mathcal{U}}$ with $\mathbf{T}$,
you get back the starting geometry $\Omega_{\mathcal{I}}$. Let
$\Xvec_{\mathcal{I}}$ denote the coordinates in $\Omega_{\mathcal{I}}$, and
let $\mathbf{d}$ denote the displacement field which is the solution
of \eqref{some_force_balance_eq} with $\mathbf{T}$ representing
the traction on the endocardium, and which depends on the reference
coordinates. What we want, is to find
$\Xvec_{\mathcal{U}}$ so that $\Xvec_{\mathcal{I}} =
\Xvec_{\mathcal{U}} + \mathbf{d}(\Xvec_{\mathcal{U}})$. To
find  $\Xvec_{\mathcal{U}}$ we let
\begin{align}
  \Xvec_{n+1} = g(\Xvec_n) =  \Xvec_{\mathcal{I}} - \mathbf{d}(\Xvec_n),
  && \Xvec_0 = \Xvec_{\mathcal{I}}.
\end{align}
According to Banach Fixed Point Theorem, the mapping $g$ has a
fixed-point if $\| \nabla g \|_{\infty} = \| \nabla \mathbf{d}
\|_{\infty} < 1$. This means that as long as the deformation
resulting from the load $\mathbf{T}$ do not result in a stretch that
is more than $100 \%$, this method will converge. Of course the
question about convergence of the fixed point method is one
thing. Another thing is the convergence of the nonlinear solver.
For bi-ventricular geometries, the backward displacement method might
fail to converge especially if the tissue is soft and the pressure in
the right ventricle is large. In this case it can happen that the
right ventricle free wall collided with the septal wall when
subtracting the displacement. An example of this is illustrated in
Figure~\ref{fig:unloading_fail}.

An alternative to the backward displacement method that do not
necessary converge towards a fixed point but provides an a more stable
way of obtainind an unloaded configuration was proposed in Ragahavan
et. al \cite{raghavan2006non}. Let $k$ be a scalar, and suppose we can write
\begin{align}
  \Xvec_{\mathcal{I}} = \Xvec_{\mathcal{U}} + k \cdot \mathbf{d}(\Xvec_{\mathcal{U}}).
\end{align}
Of course this assumption might not be true, but we can find a value of
$k$ that minimizes some residual using a 1D optimization
algorithm. Note that choosing $k = 1$, would be the same as applying
one iteration of the backward displacement method.

Since the backward displcement method really does converge towards a
fixed point, it is favorable to the Rahavan method. Therefore, a
hybrid approach is to apply the backward displacement method as
long as the non-linear solver converge, and switch to the Rahavan
method upon divergence.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{chapters/introduction/figures/unloading_fail/unloading_fail.png}
\caption{Backward displacement method might fail for bi-ventricular
  geometries. This can be seen be noticing that the RV free wall is
  buckling and colliding with the septum.}
\label{fig:unloading_fail}
\end{figure}

For a transversely isotropic material (or orthotropic), the change in
fiber field when going from one geometry to another, has to be
considered. The natural thing to do is apply a \emph{Piola
  transformation} to the fibers in the image based geometry: Let
$\F_{n}= \I - \Grad \left(\mathbf{d}(\Xvec_n) \right)$ and left
$\ef^{n} = J^{-1}\F_{n} \ef$ where $\ef$ is the fiber field on
$\Omega_{\mathcal{I}}$. Another way is to regenerate the fibers (see
Section \ref{sec:rule_based_fiber}) using the new unloaded geometry as
input. We will see in the following numerical experiment that non of
these methods gives the desired result.

% \begin{remark}
%   The author has observed that when applying the backward
%   displacement method to bi-ventricular geometries subjected to a high
%   pressure in the right ventricle, and/or a thin RV free wall, then
%   the right ventricular free wall will in some cases collide with the
%   septum. Figure \ref{fig:unloading_fail} shows an example of
%   phenomena. In these cases a more robust approach (but less accurate)
%   is let $\Xvec_{n+1} = \Xvec_{\mathcal{I}} - k \mathbf{d}(\Xvec_n)$,
%   where $\Xvec_n$ is the results from the last converged fixed point
%   iteration, and use an optimization algorithm to determine the value
%   of $k$ than minimizes some residual. A similar approach has been
%   proposed in \cite{raghavan2006non}. 
% \end{remark}

@@

The other approach, called variational data assimilation, also known as
4D-var, is based on minimizing a cost functional that
represents the mismatch between observations and simulations, with
possibly additional regularization terms added to the objective
functional. We will focus this approach since this is the approach
that is taken in this thesis. For a more thourough review of other
methods we refer to \cite{chapelle2013fundamental}.

% The general setup for a variational data assimilation problem is the
% following.
\subsubsection{Pipeline}
\label{sec:data_assimilation_pipeline}

We will now explain the general setup and the main steps in variational data
assimilation. In order to keep consistent notation we will refer to
the \emph{state variable} as $\mathbf{w}$, which is our case
represents jointly the displacement and hydrostatic pressure
$\mathbf{w}=(\mathbf{u}, p)$.

The state varibles are described by some \emph{phyiscal forward
model} $\mathcal{M}$, which is our case is governed by the force balance
equation in \eqref{eq:force_balance_strong}. Moreover this model
typically depends on some parameters $\mu$ that we want to
adjust based on the observation at hand. For example, the unknown
parameters could be related to the stiffness of the myocardium which
which we want to adjust to each individual patient based on some
clinical obervations. We can therefore assume that the underlying
model can be written in the form $\mathcal{M}(\mathbf{w}, \mu) = 0$.

We are also given a set of measurements (or observations) that we want
to assimilate, and we denote a single observation by $y$.
For example, $y$ might by a volume measurements at some given time
points in the cardiac cycle, measuered using e.g unltrasound and image
processing techniques. The \emph{observation operator} $\mathcal{H}$,
is approximation of the observation and act as an operator from the
state space to the observation space. For a single obervation, we
therefore have the relation
\begin{align}
  y = \mathcal{H}(\mathbf{w}) + \xi, 
\end{align}
where $\xi = \xi_{\mathbf{\mathcal{H}}} + \xi_{\mathbf{y}}$ represents
both the error in the measurements ($\xi_{\mathbf{y}}$) (e.g
noise in the data) and error in the representation of the observation
($\xi_{\mathbf{\mathcal{H}}}$) (that $\mathcal{H}$ do not represent
the data good enough).

Finally all observations are combined into a single cost function
$\mathcal{J} = \mathcal{J}(\state, \mu)$ which represents the overall mismatch between
observations and simulations. The aim is to minimize this functional
while ensuring that the governing force balance equation is
satisfied. This can be done in the following steps

\begin{enumerate}
  \item For some initial guess $\mu_0$, solve the forward model
    $\mathcal{M} = 0$ to obtain $\state_0$.
  \item Compute $\mathcal{J}(\state_0, \mu_0)$.
  \item Compute the gradient $D \mathcal{J}(\state_0, \mu_0)$ to find
    the direction of the steepest descent.
  \item Move along the steepest direction and update the initial guess.
\end{enumerate}
The above steps are continued until we have reached a minimum for the
cost functional, e.g when the norm of the gradient becomes zero. In
practice we continue these steps until either the norm of the
gradient is below a certain tolerance or we have exceeded a maximum
number of iterations. Such a minimum may not allways exists, or there
may be several local minima. In the latter case, the choice of intial
guess determines which local minima you will end up in. The data
assimilation pipeline is summarized in
Figure~\ref{fig:data_assimilation}.

\begin{remark}
  There exists gradient-free optimization methods that can be used to
  avoid computing the gradient. Examples include genetic algorithms or
  particle swarm algorithms. The drawback with such methods is that
  they typically requires a lot of functional evaluations. In our
  case this is very computatinally exspensive, because every functional
  evaluation require us to solve the non-linear forward model. 
\end{remark}

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/data_assimilation.eps}
\caption{The different componenets involved in the data assimilation
  procedure. The mathematical description of each component is
  displayed below each box. From a patients heart we typicall extract information
  that are used as input to the model. This can for instance be
  information about geometry or information about boundary
  conditions. This model input is used to create a mathematical model
  of the heart, which also depends on some parameters $\mu$.
  These unknown parameters are determined by minimizing the distance
  between the clinical obervations $y$ and the modeled observation
  $\mathcal{H}(\mathbf{w})$.}
\label{fig:data_assimilation}
\end{figure}

We will now describe some of the main observation operators used in
this thesis before we explain how to solve the data assimilation
problem efficiently using the adjoint method.
